From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-08 Pi+/paper/equivalence.tex
Revise some things in the appendix

As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal strucuture. Since each object is a
natural number generated by $\zero$ and $\suc$, this makes $\PiHatCat$ similar to a \emph{PROP}, that is, a products and
permutations category.

@@

\section{Normalisation of reversible circuits}
\label{sec:applications}

Using our semantics, we can normalise a circuit writen in $\PiLang$ -- we first translate it to $\PiHatLang$, then
compute a permutation, and quote it back to $\PiPlusLang$. Note that the normalisation happens in going from
$\PiHatLang$ to $\UFin$ and back to $\PiHatLang$.

\begin{definition}
  \begin{gather*}
    \begin{aligned}
      \normt_{0} & : U \to \UPlus                                            \\
      \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0}
    \end{aligned}
    \qquad
    \begin{aligned}
      \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
      \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
    \end{aligned}
  \end{gather*}
\end{definition}

We revisit the examples in~\Cref{sec:examples} but instead of working with circuits operating on sequences of bits, we
work with circuits operating on algebraic data types in $\PiLang$.

\paragraph*{Normalisation by evaluation.} Consider the following reversible function: $\Afun{RESET}(b,b_1,\ldots,b_n) =
(b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is logical-or and $\underline{\vee}$ is
exclusive-or. The function sets or resets the first bit depending on whether any of the other bits is true or not. The
natural definition for this function is recursive: modulo some re-shuffling of the bits, the strategy is to examine the
bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and terminate; otherwise we continue making
recursive calls until we reach the last bit at which point we return the incoming value with no change:

@@

\resettwo{}

\noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the introduction.
Our Agda infrastructure provides tools to normalise all programs to their normal form. For \Afun{reset 2} we get:

\resetnormtwo{}

In principle, the normalised program can be produced following two strategies: (i) by repeatedly applying the rewrite
rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
using a normalisation-by-evaluation (NbE) process that evaluates the program to a permutation on a finite set of 8
elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
left is compiled to the sequence of four adjacent transpositions on the right:

@@

\paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
permutations. Indeed, instead of writing a program for \Afun{reset 2}, one could simply specify the desired permutation
as:

\resetperm{}

\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Simply
``quoting'' this permutation generates the same normalised program \Afun{reset2Norm} above.

\paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
following program:

\adder{}

\noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
\verb|reversibleOr1| from the introduction.

>>>>

As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure. Since each object is a
natural number generated by $\zero$ and $\suc$, this makes $\PiHatCat$ similar to a \emph{PROP}, that is, a products and
permutations category.

@@

\section{Normalisation of Reversible Circuits}
\label{sec:applications}

Using our semantics, we can normalise, synthesise, prove equivalence, and generally reason about~$\PiLang$
programs. The two key definitions are presented below.

\begin{definition}[Normalisation of $\PiLang$ programs]
  \begin{gather*}
    \begin{aligned}
      \normt_{0} & : U \to \UPlus                                            \\
      \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0}
    \end{aligned}
    \qquad
    \begin{aligned}
      \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
      \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
    \end{aligned}
  \end{gather*}
\end{definition}

\noindent Normalisation consists of translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting
back to $\PiPlusLang$. Note that the normalisation happens in the step from $\PiHatLang$ to $\UFin$ and back to
$\PiHatLang$ and that normalisation also provides a decision procedure for program equivalence. Synthesis consists of
quoting permutations. More general, user-guided, reasoning can be done using the sound and complete level-2
combinators to rewrite $\PiLang$ programs.

Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing it, and the
corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal for both circuit is
computed to be, establishing their equivalence:

\medskip
\resetnormtwo{}

Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is also possible
to simply quote the desired permutation:

\medskip
\resetperm{}

\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Quoting this
permutation generates the same normalised program which can then be composed with a map from $\PiLang$ to
$\PiPlusLang$ to produce a program matching the desired structured types.

% \paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
% permutations. Indeed, instead of writing a program for \Afun{reset 2}, one could simply specify the desired permutation
% as:

@@

%% \resettwo{}

% \noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
% provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
% reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the introduction.

% In principle, the normalised program can be produced following two strategies: (i) by repeatedly applying the rewrite
% rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
% using a NbE process that evaluates the program to a permutation on a finite set of 8
% elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
% permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
% left is compiled to the sequence of four adjacent transpositions on the right:

@@

% \paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
% special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
% one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
% following program:

% \adder{}

% \noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
% same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
% \verb|reversibleOr1| from the introduction.
